---
title: "ESS 330 Final Project: Emissions in the Top CO₂ Polluting Countries Before and After COVID-19"
author:
  - name: "Yazeed Aljohani"
    affiliation: "Colorado State University"
  - name: "Josh Puyear"
    affiliation: "Colorado State University"
  - name: "Cade Vanek"
    affiliation: "Colorado State University"
date: 2025-04-24"
bibliography: references.bib
keywords: [CO2 emissions, COVID-19, GHG, workflow, ANOVA, sustainability]
abstract: |
  This project investigates trends in per capita greenhouse gas (GHG) emissions from the five largest CO2-polluting countries before, during, and after the COVID-19 pandemic. Using `ghg_excluding_lucf_per_capita` as the main [response?] variable, we analyze patterns in emissions and identify key predictors across sectors (e.g., manufacturing, transport). The study applies workflow modeling and ANOVA to examine emission changes and variation across countries. Results will inform strategies to reduce emissions and guide sustainable policy efforts.
format:
  html:
    code-fold: true
---

# Introduction, Background, and Motivation

Climate change remains one of the most urgent global challenges, driven in large part by human greenhouse gas emissions. Among these, carbon dioxide is the most significant. Carbon dioxide is released through many activities such as burning fossil fuels, industrial production, and land-use practices. Even though international efforts like the Paris Agreement have pushed countries to reduce emissions, many continue to show increasing trends. A small group of countries contributes the most emissions globally, both in total and per person. These major contributors include countries like China, the United States, and India [@friedlingstein2019global].

This project focuses on the five largest carbon dioxide polluting countries: China, the United States, India, Russia, and Japan. These countries are responsible for a significant share of global emissions and reflect a range of economic systems, energy sources, and industrial structures. To allow meaningful comparisons, we center our analysis on greenhouse gas emissions per person, using the variable ghg_excluding_lucf_per_capita. This variable removes land-use change from the emissions total, isolating emissions from industrial activity and fossil fuel use. By using this approach, we can better understand how much pollution is tied to each person’s activity and consumption within each country.

The COVID-19 pandemic created a global disruption that affected travel, trade, and production. Lockdowns and other restrictions slowed down factories, reduced transportation, and caused shifts in energy use. As a result, global carbon dioxide emissions in 2020 dropped by approximately 5.4 percent, marking the largest single-year decrease in history [@forster2020current]. This sudden drop was mostly due to temporary changes. As countries began reopening in 2021 and 2022, emissions started climbing again.

Our goal is to analyze how emissions changed in these five countries before, during, and after the pandemic. We look at three main periods: before COVID from 2015 to 2019, the year COVID began in 2020, and the early recovery years from 2021 to 2022. By studying per person emissions during these periods, we can see whether there were real shifts in emissions behavior and whether those shifts continued into recovery. We also want to understand what kinds of emissions changed the most. For example, did transportation-related emissions drop more than industrial ones? Did countries with manufacturing-based economies recover differently than those with service-based ones?

To explore these questions, we use two main methods. First, we build a workflow model using the tidymodels package in R. This lets us see which emission sources are the strongest predictors of overall greenhouse gas output per person. For example, we compare emissions from coal, oil, gas, cement, and other industry sectors. We also include factors like population and economic output. By analyzing these predictors, we can understand which sources matter most in each country and how they are connected to changes over time. Second, we use ANOVA to test whether the changes in emissions across the three time periods are statistically meaningful. We run this analysis for each country and compare their patterns.

The reason for focusing on per person emissions is that total emissions alone do not tell the full story. Countries with large populations may seem like major polluters in total, but their emissions per person might be lower than smaller, high-emitting countries. A per person view also reflects differences in daily life, energy use, and infrastructure. This is important for climate policy, because it shows whether emissions are driven by large populations, high-intensity consumption, or both.

Our analysis helps link data analysis to climate solutions. By identifying what types of emissions dropped the most during the pandemic and which ones came back quickly, we can suggest where to focus efforts. If, for instance, oil-related emissions dropped during lockdowns but then returned quickly, that might show a need for deeper investment in alternative transportation options. If manufacturing emissions remained stable, that could show how industrial processes are harder to change without long-term plans and policies.

This project is also a chance to make emissions data more relatable. People often hear that emissions went down during the pandemic, but they may not know what caused the change or how long it lasted. By using clear visuals and country comparisons, we can help explain these trends to wider audiences. This is especially important as the world faces increasing climate-related disasters. Understanding emission trends can support smarter planning and better communication around sustainable transitions.

Lastly, we are motivated by a bigger question: what can we learn from this accidental global experiment? The pandemic gave the world a chance to see what happens when daily life suddenly changes. Some of those changes, like working from home or using less transportation, might become long-term habits. Others may disappear as economies reopen. By analyzing which emission reductions lasted and which faded, we can start to imagine more realistic and effective ways to slow down climate change.

Add to the introduction(Josh):

Energy-related emissions are an important aspect of analyzing CO2 emission changes before and after COVID because demands for energy will continue. Depending on where the five selected countries obtain fuels for energy generation, energy use plays a role in predicting total CO2 emissions for each country and can present opportunities for reductions in emissions. CO2  emitted per kilowatt-hour can provide insight into how much CO2 can be attributed to energy use.

Questions (Josh):

Were government economic and public restrictions the same or different during the lockdown? How does this relate to changing CO2 emissions per capita over time for the five selected countries?

What is the share of CO2  emissions from the top five emitters post COVID?

Which variables explain CO2  emissions per kilowatt-hour for the top five emitters?

# **Data Exploration**

```{r}
library(tidyverse)
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)

data <- read_csv("data/owid-co2-data (1).csv")

meta <- read_csv("data/owid-co2-codebook (1).csv")

```
## Narrowing Data to 2015-2022
```{r}
data_filtered <- data |>
  filter(year >= 2015, year <= 2022) |>
  filter(!is.na(iso_code) & nchar(iso_code) == 3)
```

```{r}
# Getting top 5 CO2 emitting countries by total GHG (excluding land use)- This is total cumulative ghg emissions

top_emitters <- data_filtered |>
  group_by(country) |>
  summarise(total_ghg = sum(total_ghg_excluding_lucf, na.rm = TRUE)) |>
  arrange(desc(total_ghg)) |>
  slice_head(n = 5) |>
  pull(country)

# Filtering data to include only those countries and select relevant variables
df <- data_filtered |>
  filter(country %in% top_emitters) |>
  select(country, year, ghg_excluding_lucf_per_capita,
         coal_co2, gas_co2, oil_co2, cement_co2, other_industry_co2)

```

```{r}
# Adding period category (pre, during, post COVID)
df <- df |>
  mutate(period = case_when(
    year <= 2019 ~ "pre_covid",
    year == 2020 ~ "during_covid",
    year >= 2021 ~ "post_covid"
  )) |>
  mutate(period = factor(period, levels = c("pre_covid", "during_covid", "post_covid")))

```

Per capita ghg emissions top 5 polluters
```{r}
ggplot(df, aes(x = factor(year), y = ghg_excluding_lucf_per_capita, color = country, group = country)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Per Capita GHG Emissions (Excl. Land Use) by Country (2015–2022)",
    x = "Year", y = "GHG per Capita (tCO₂e)"
  ) +
  theme_minimal()

```

```{r}
df_long <- df |>
  pivot_longer(cols = c(coal_co2, gas_co2, oil_co2, cement_co2, other_industry_co2),
               names_to = "sector", values_to = "emissions")

ggplot(df_long, aes(x = factor(year), y = emissions, color = sector, group = interaction(sector, country))) +
  geom_line() +
  geom_point() +
  facet_wrap(~ country) +
  labs(title = "Sector-specific CO₂ Emissions (2015–2022)", x = "Year", y = "Emissions (MtCO₂)") +
  theme_minimal()

```

```{r}
# Running ANOVA on each country if they have more than one period
for (c in unique(df$country)) {
  df_country <- df |> filter(country == c)
  if (n_distinct(df_country$period) > 1) {
    cat("\n--- ANOVA for", c, "---\n")
    model <- aov(ghg_excluding_lucf_per_capita ~ period, data = df_country)
    print(summary(model))
  } else {
    cat("\n--- Skipped ANOVA for", c, ": not enough periods ---\n")
  }
}

```

#### Data Description

We used data from Our World in Data’s greenhouse gas emissions dataset, which compiles country-level data from sources like the Global Carbon Project and BP Statistical Review [@owid-scaling-up-ai]. The dataset covers emissions from fossil fuels, industry, and land-use change.

For this project, we focused on the years 2015 to 2022 and selected the top five CO₂-polluting countries: China, the United States, India, Russia, and Japan. We used variables such as:

-   ghg_excluding_lucf_per_capita (our main variable),

-   Sector-based emissions: coal_co2, oil_co2, gas_co2, cement_co2, other_industry_co2.

#### Data Cleaning

We filtered the dataset to include:

-   Only rows with valid ISO country codes

-   Only years 2015–2022

-   The top 5 countries based on total GHG emissions (excluding land-use change)

We also created a period variable to group data into pre-COVID (2015–2019), during COVID (2020), and post-COVID (2021–2022). This helped us compare emissions trends across time. Minimal cleaning was needed beyond this filtering and labeling.

# Methods (Josh)

  Our research focuses on revealing relationships between countries and between predictor variables of greenhouse emissions. The dataset from Oxford's Our World in Data includes emissions levels from the industrial revolution up to 2023 and is a robust dataset. The dataset is publicly posted to https://ourworldindata.org/co2-and-greenhouse-gas-emissions. To compare CO2 emissions before, during, and after the pandemic, we will be using the tidymodels package in R along with dplyr. We will make this comparison by comparing transportation and industrial emissions, then comparing total emissions from five key predictor categories including coal, oil, gas, cement and other. Our main response will be ghg_excluding_lucf_per_capita, but other measures of CO2 will also be explored. We will be testing predictor variables with linear regression, neural network, random forest, decision tree, and boost tree models to find predictor variables.

  Restrictions to public and economic activity during the 2020-2021 COVID lockdown period prevented people from leaving their houses to consume goods and do business, activities that would under normal circumstances indirectly result in more greenhouse gas emissions. This study refers to The Oxford COVID-19 Government Response Tracker (OxCGRT) for public and economic restriction indices in the five focus countries. Differences in average restriction index for each country were tested with ANOVA. We used the ggplot2 package to visualize COVID-19 restrictions as well as visualizing data from the CO2 and Greenhouse Gas emissions data by Ritchie, Rosado and Roser. 
  The proportion of per capita emissions from the top five cumulative CO2  emitting countries to the full record of countries during the 2015-2022 period revealed the impact that these countries can have to curb CO2 emissions in the future. Then, we used CO2 per unit energy emissions in 2015-2022 to illustrate how energy, a constant necessity, contributes to the carbon footprint of these countries. The predictor variables gdp_percap, gas_co2_per_capita, share_global_coal_co2, coal_co2_per_capita, cumulative_luc_co2, oil_co2_per_capita, and share_global_luc_co2 (Table 1) were chosen to model CO2 emissions per unit energy. The ggplot packages rsample, parsnip, baguette, and workflowsets were used to test multiple models over the dataset to build a regression model. The best model was analyzed with the vip package to reveal the most explanatory variables that predicted CO2 emissions per unit energy from the selected variables.

**Table 1.** Predictor variables for CO2 emitted per kilowatt-hour
```{r, echo = TRUE}
meta %>% 
  filter(column %in% c("gdp_percap", "gas_co2_per_capita", "share_global_coal_co2", "coal_co2_per_capita", "cumulative_luc_co2", "oil_co2_per_capita", "share_global_luc_co2")) %>% 
  print()

```


```{r, echo = FALSE}
#include a table of variables and their meaning (reference df above)
# copy metadata from orig proj to here
#names(df)
```
## Understanding Public Restrictions

```{r, echo = FALSE}
library(readr)
library(readxl)
```


```{r, echo = FALSE}
#loading all dataset
govres <- read_excel("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/Gov_Responses2Covid19_last.xlsx", sheet = "Dataset")

govresdes<- read_excel("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/Gov_Responses2Covid19_last.xlsx", sheet = "Description")

govres_econ <- read_excel("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/Gov_Responses_Sources.xlsx", sheet = "Econ sources")

govres_pubhealth <- read_excel("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/Gov_Responses_Sources.xlsx", sheet = "Public health sources")

#View(govres)
#View(govres_source)
#view(govresdes)

```

```{r, echo = FALSE}
#names(govres)
#summary(govres)
```


```{r, echo = TRUE}
# Getting top 5 CO2 emitting countries by total GHG (excluding land use)

#filter to top 5 countries

#need to actually just select for the five most ghg emitting countries
top_emitters_res <- govres |>
  group_by(country) |>
  filter(country %in% c("China", "India", "Russia", "United States of America", "Japan"))

# Filtering data to include only those countries and select relevant variables
df2 <- top_emitters_res %>% 
  select(country, date, Rigidity_Public_Health, Economic_Measures) %>% 
  mutate(Public = as.numeric(Rigidity_Public_Health)) %>% 
  mutate(Economic = as.numeric(Economic_Measures)) %>% 
  select(!c(Rigidity_Public_Health, Economic_Measures))
```

```{r, echo = TRUE}
df2_long <- df2 |>
  pivot_longer(cols = c(Public, Economic),
               names_to = "Restriction", values_to = "Rigidity") %>% 
  mutate(Date = as.Date(date))

#making a faceted plot
ggplot(df2_long, aes(x = Date, y = Rigidity, color = Restriction, group = interaction(Restriction, country))) +
    theme(
    axis.text.x = element_text(size = 6, angle = 60, hjust = 1),  # Remove extra comma here
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_x_date(date_labels = "%b %Y", 
               date_breaks = "3 months") +
  scale_y_continuous(limits = c(0, 1)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ country) +
  labs(title = "Restrictions During COVID's Lockdown", x = "Date", y = "Restriction Index") +
  theme_minimal()


ggsave("countryrestrict.pdf", plot = last_plot(), path = "C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/imgs", width = 12, height = 6)
```
**Figure 1.** Daily COVID lockdown restriction indices (Oxford, 2022), separated by Economic and Public.

**Public Restriction Indices**
There is a statistically significant difference (p < 0.000) in both economic and public restriction index for the top 5 biggest cumulative CO2 emitters (Figure 1). The United States has significantly lower public restrictions (p < 0.000) than every other country except China, where its public restrictions are similar (p = .3660).


```{r, echo = FALSE}
library(car)

#shapiro test for Public restrictions
normspub <- df2 |> 
  nest(data = -country) |>
  mutate(
    Shapiro = map(data, ~ shapiro.test(na.omit(.x$Public))),  # Remove NAs before applying shapiro.test
    n = map_dbl(data, nrow),
    glance_shapiro = map(Shapiro, broom::glance)
  ) |>
  unnest(glance_shapiro)

flextable::flextable(dplyr::select(normspub, country, n, statistic, p.value)) |>
  flextable::set_caption("Shapiro-Wilk normality test for restrictions in each country")
#the public restriction dates of lockdown are not normally distributed

#testing economic restrictions
normsec <- df2 |> 
  nest(data = -country) |>
  mutate(
    Shapiro = map(data, ~ shapiro.test(na.omit(.x$Economic))),  # Remove NAs before applying shapiro.test
    n = map_dbl(data, nrow),
    glance_shapiro = map(Shapiro, broom::glance)
  ) |>
  unnest(glance_shapiro)

flextable::flextable(dplyr::select(normsec, country, n, statistic, p.value)) |>
  flextable::set_caption("Shapiro-Wilk normality test for restrictions in each country")
#economic restriction dates aren't normally distributed either
```



```{r, echo = FALSE}
#testing for statistical significance in differences between indices for how countries handled the lockdown

kruskal.test(Public ~ country, data = df2)

kruskal.test(Economic ~ country, data = df2)

```

post-hoc test to see which ones are different

```{r, echo = FALSE}
# using Dunn's test to see which variables are different
library(dunn.test)
# Perform Dunn's Test for pairwise comparisons
dunn.test(df2$Public, g = df2$country)

dunn.test(df2$Economic, g = df2$country)
```
Results:
The United States has significantly lower public restrictions (p < 0.000) than every other country except China.

Proportion of CO2 emitted by top 5 countries in 2022

```{r, echo = FALSE}
#for all greenhouse gases
total_emissions <- data %>% 
  select(total_ghg, year, country, iso_code) %>%
  filter(!is.na(iso_code)) %>% 
  group_by(country, na.rm = TRUE) %>% 
  filter(year == 2022) %>% 
  ungroup()

#in the five most polluting countries
totem1 <- total_emissions %>% 
  filter(country %in% c("United States", "China", "India", "Japan", "Russia")) %>% 
  summarize(sum(total_ghg))

#in the rest of the world
totem2 <- total_emissions %>% 
  filter(!country %in% c("United States", "China", "India", "Japan", "Russia")) %>% 
  summarize(sum(total_ghg, na.rm = TRUE))

27168.36 / (27168.36 + 26138.06)

```

```{r, echo = FALSE}
#for co2_excluding_luc_per_capita
percap_emissions <- data %>% 
  select(ghg_excluding_lucf_per_capita, year, country, iso_code) %>%
  filter(!is.na(iso_code)) %>% 
  group_by(country, na.rm = TRUE) %>% 
  filter(year == 2022) %>% 
  ungroup()

#in the five most polluting countries
totemcap1 <- percap_emissions %>% 
  filter(country %in% c("United States", "China", "India", "Japan", "Russia")) %>% 
  summarize(mean(ghg_excluding_lucf_per_capita))

#in the rest of the world
totemcap2 <- percap_emissions %>% 
  filter(!country %in% c("United States", "China", "India", "Japan", "Russia")) %>% 
  summarize(mean(ghg_excluding_lucf_per_capita, na.rm = TRUE))
```

Results
The top five biggest co2 emitted of the world emitted a total of 27168.36 million metric tons of greenhouse gases in 2022. That year, the other 213 countries emitted metric tons of greenhouse gasses, meaning 50.97 percent of greenhouse gas emissions come from the top 5 most polluting countries. The mean emissions per capita in the top 5 biggest greenhouse gas emitters was 10.45 CO2 equivalents versus a mean of only 5.59 CO2 equivalents for the rest of the countries.

## CO2 per Unit Energy Modeling

```{r, echo = TRUE}
#co2_per_unit_energy
 
df3 <- data_filtered %>% 
  group_by(country) %>% 
  filter(country %in% c("China", "India", "Russia", "United States", "Japan")) %>%
  select(country, year, co2_per_unit_energy,
         ghg_excluding_lucf_per_capita) %>% 
  mutate(percap = ghg_excluding_lucf_per_capita, energy = co2_per_unit_energy)


ggplot(df3, aes(x = factor(year), y = co2_per_unit_energy, color = country, group = country)) +
  geom_line() +
  geom_point() +
  labs(
    title = " 2015 - 2022 CO2 per kilowatt-hour emissions",,
    x = "Year", y = "CO2 produced per kwh emissions"
  ) +
  theme_minimal()

```
**Figure 2.** CO2 per kilowatt-hour emissions for the top five cumulative emitters.

**Proportion of CO2  emitted by the five selected countries**
In 2022, India and Russia’s CO2 emissions per kilowatt-hour have risen, while China, the United States, and Japan’s emissions have decreased. (Figure 2.) The top five biggest co2 emitted of the world emitted a total of 27168.36 million metric tons of greenhouse gases in 2022. That year, the other 213 countries emitted metric tons of greenhouse gasses, meaning 50.97 percent of greenhouse gas emissions come from the top 5 most polluting countries. The mean emissions per capita in the top 5 biggest greenhouse gas emitters was 10.45 CO2 equivalents versus a mean of only 5.59 CO2 equivalents for the rest of the countries.


```{r, echo = FALSE}
#shapiro test for Public restrictions
normspub2 <- df3 |> 
  nest(data = -country) |>
  mutate(
    Shapiro = map(data, ~ shapiro.test(na.omit(.x$energy))),  # Remove NAs before applying shapiro.test
    n = map_dbl(data, nrow),
    glance_shapiro = map(Shapiro, broom::glance)
  ) |>
  unnest(glance_shapiro)

flextable::flextable(dplyr::select(normspub2, country, n, statistic, p.value)) |>
  flextable::set_caption("Shapiro-Wilk normality test for energy consumption in each country")

#This table is giving normal distribution for the p-values, so we'll do ANOVA

```

ANOVA for co2 emissions per unit energy

```{r, echo = FALSE}
library(ggpubr)

res_aov <- aov(formula = co2_per_unit_energy ~ country, data = df3)

#making sure residuals have a normal distribution
gghistogram(res_aov$residuals)

shapiro.test(res_aov$residuals)
#p-value is .8234, so the data fit a normal distribution
```


```{r, echo = FALSE}
library(ggpubr)
norm_usa <- df3 %>% group_by(country) %>% 
filter(country == "United States")

gghistogram(norm_usa$energy)

shapiro.test(df3$energy)

#it was strange how high the p-values were for the shapiro.test so I wanted to make a histogram. The USA histogram doesn't appear to have a normal distribution

```


p-values from ANOVA

**Table 2.** Summary statistics of an ANOVA analysis of CO2 emitted per kilowatt-hour for the top 5 cumulative CO2 emitters.
```{r, echo = TRUE}
summary(aov(formula = co2_per_unit_energy ~ country, data = df3))
```
**Difference in CO2  per kilowatt-hour emissions between countries**
There is a statistically significant difference (p < 0.000) in CO2 per kilowatt-hour for the top 5 biggest cumulative CO2 emitters (Table 2). 


## Predicting CO2 Emissions per unit Energy

Mutate data, check for Interaction terms
``` {r, echo = TRUE}
#set a seed
set.seed(341)

#doing a correlation test for variables
1
ghg_per_eny <- data_filtered %>% 
  mutate(period = case_when(
    year <= 2019 ~ "pre_covid",
    year == 2020 ~ "during_covid",
    year >= 2021 ~ "post_covid"
  )) %>% 
  group_by(country) %>% 
  mutate(gdp_percap = gdp/population) %>% 
  ungroup() %>% 
select(co2_per_unit_energy, share_global_luc_co2,
gas_co2_per_capita, oil_co2_per_capita, gdp_percap, coal_co2_per_capita, share_global_coal_co2, cumulative_luc_co2, share_global_luc_co2) %>%
  drop_na

# cor(ghg_per_eny)

```
None of the variables had strong enough correlations to make interaction terms.

## making testing and training data for energy consumption

```{r, echo = TRUE}
#find recipe format from lab 6 / model daily assignments
library(rsample)

ghg_per_eny_split <- initial_split(ghg_per_eny, prop = .8)
ghg_per_eny_train <- training(ghg_per_eny_split)
ghg_per_eny_test  <- testing(ghg_per_eny_split)


ghg_per_eny_cv <- vfold_cv(ghg_per_eny, v = 10)
```


```{r, echo = TRUE}
#attempted recipe format
library(recipes)

rec_energy <- recipe(co2_per_unit_energy ~
gas_co2_per_capita + oil_co2_per_capita + gdp_percap + coal_co2_per_capita + share_global_coal_co2 + cumulative_luc_co2 + share_global_luc_co2, data = ghg_per_eny) %>%
  step_naomit(all_predictors(), all_outcomes())

#ok, it's not worth logging any of these predictors because the models I'm making don't require normal distributions
```

Making different models to test against
```{r, echo = TRUE}
#making the models
library(parsnip)

boost <- boost_tree() %>%
  # define the engine
  set_engine("xgboost") %>%
  # define the mode
  set_mode("regression")

nnet <- bag_mlp() %>%
  # define the engine
  set_engine("nnet") %>%
  # define the mode
  set_mode("regression")

dtree <- decision_tree() %>%
  # define the engine
  set_engine("rpart") %>%
  # define the mode
  set_mode("regression")

rf <- rand_forest(
  mtry = 5,
  trees = 1000,
  min_n = 5
) %>%
  set_engine("ranger", importance = "impurity") %>%  # <-- ADD THIS
  set_mode("regression")
```

Plotting the best predictive models of emissions per unit energy
```{r, echo = TRUE}
library(workflowsets)
library(baguette)

wf <-  workflow_set(list(rec_energy), 
                  list(boost,
                       nnet,
                       dtree,
                       rf)) %>%
  workflow_map('fit_resamples', resamples = ghg_per_eny_cv) 

autoplot(wf)
```
**Figure 4.** Ranking different models' root mean squared error and R-squared error values, tested against CO2 per kilowatt hour emissions data for the top five cumulative CO2 emitters.

Results:

The graph shows that random forest is the best model for predicting carbon emissions per unit energy.


**Table 3.** Ranking Model Testing Results
```{r, echo = TRUE}
rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```

The random forest does best in this situation to model ghg emissions per unit energy

### Making a workflow to predict ghg emitted per kilowatt-hour
```{r, echo = TRUE}
library(tidymodels)
rf_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec_energy) %>%
  # Add the model
  add_model(rf) %>%
  # Fit the model to the training data
  fit(data = ghg_per_eny_train) 

rf_data <- augment(rf_wf, new_data = ghg_per_eny_test)

dim(rf_data)

```

finding the most statistically significant predictors
```{r, echo = TRUE}
rf_model <- extract_fit_engine(rf_wf)
vip::vip(rf_model)
```
**Figure 5.** Ranking importance of each predictor variable at explaining CO2 emissions per kilowatt-hour in the random forest model selected for analysis.

**Predictor variables for CO2  per kilowatt-hour**
The graph shows that random forest is the best model for predicting carbon emissions per unit energy, with an R-squared of 0.91. (Figure 4). The most explanatory variable for predicting CO2 emissions per unit of energy consumption during the the 2015 to 2022 period in the five countries examined was GDP per capita followed by CO2 emitted from gas per capita (Figure 5).
