---
title: "ESS 330 Final Project: Emissions in the Top CO₂ Polluting Countries Before and After COVID-19"
author:
  - name: "Yazeed Aljohani"
    affiliation: "Colorado State University"
  - name: "Josh Puyear"
    affiliation: "Colorado State University"
  - name: "Cade Vanek"
    affiliation: "Colorado State University"
date: 2025-04-24"
bibliography: references.bib
keywords: [CO2 emissions, COVID-19, GHG, workflow, ANOVA, sustainability]
abstract: |
  This project investigates trends in per capita greenhouse gas (GHG) emissions from the five largest CO2-polluting countries before, during, and after the COVID-19 pandemic. Using `ghg_excluding_lucf_per_capita` as the main [response?] variable, we analyze patterns in emissions and identify key predictors across sectors (e.g., manufacturing, transport). The study applies workflow modeling and ANOVA to examine emission changes and variation across countries. Results will inform strategies to reduce emissions and guide sustainable policy efforts.
---

# Introduction, Background, and Motivation

Climate change remains one of the most urgent global challenges, driven in large part by human greenhouse gas emissions. Among these, carbon dioxide is the most significant. Carbon dioxide is released through many activities such as burning fossil fuels, industrial production, and land-use practices. Even though international efforts like the Paris Agreement have pushed countries to reduce emissions, many continue to show increasing trends. A small group of countries contributes the most emissions globally, both in total and per person. These major contributors include countries like China, the United States, and India [@friedlingstein2019global].

This project focuses on the five largest carbon dioxide polluting countries: China, the United States, India, Russia, and Japan. These countries are responsible for a significant share of global emissions and reflect a range of economic systems, energy sources, and industrial structures. To allow meaningful comparisons, we center our analysis on greenhouse gas emissions per person, using the variable ghg_excluding_lucf_per_capita. This variable removes land-use change from the emissions total, isolating emissions from industrial activity and fossil fuel use. By using this approach, we can better understand how much pollution is tied to each person’s activity and consumption within each country.

The COVID-19 pandemic created a global disruption that affected travel, trade, and production. Lockdowns and other restrictions slowed down factories, reduced transportation, and caused shifts in energy use. As a result, global carbon dioxide emissions in 2020 dropped by approximately 5.4 percent, marking the largest single-year decrease in history [@forster2020current]. This sudden drop was mostly due to temporary changes. As countries began reopening in 2021 and 2022, emissions started climbing again.

Our goal is to analyze how emissions changed in these five countries before, during, and after the pandemic. We look at three main periods: before COVID from 2015 to 2019, the year COVID began in 2020, and the early recovery years from 2021 to 2022. By studying per person emissions during these periods, we can see whether there were real shifts in emissions behavior and whether those shifts continued into recovery. We also want to understand what kinds of emissions changed the most. For example, did transportation-related emissions drop more than industrial ones? Did countries with manufacturing-based economies recover differently than those with service-based ones?

To explore these questions, we use two main methods. First, we build a workflow model using the tidymodels package in R. This lets us see which emission sources are the strongest predictors of overall greenhouse gas output per person. For example, we compare emissions from coal, oil, gas, cement, and other industry sectors. We also include factors like population and economic output. By analyzing these predictors, we can understand which sources matter most in each country and how they are connected to changes over time. Second, we use ANOVA to test whether the changes in emissions across the three time periods are statistically meaningful. We run this analysis for each country and compare their patterns.

The reason for focusing on per person emissions is that total emissions alone do not tell the full story. Countries with large populations may seem like major polluters in total, but their emissions per person might be lower than smaller, high-emitting countries. A per person view also reflects differences in daily life, energy use, and infrastructure. This is important for climate policy, because it shows whether emissions are driven by large populations, high-intensity consumption, or both.

Our analysis helps link data analysis to climate solutions. By identifying what types of emissions dropped the most during the pandemic and which ones came back quickly, we can suggest where to focus efforts. If, for instance, oil-related emissions dropped during lockdowns but then returned quickly, that might show a need for deeper investment in alternative transportation options. If manufacturing emissions remained stable, that could show how industrial processes are harder to change without long-term plans and policies.

This project is also a chance to make emissions data more relatable. People often hear that emissions went down during the pandemic, but they may not know what caused the change or how long it lasted. By using clear visuals and country comparisons, we can help explain these trends to wider audiences. This is especially important as the world faces increasing climate-related disasters. Understanding emission trends can support smarter planning and better communication around sustainable transitions.

Lastly, we are motivated by a bigger question: what can we learn from this accidental global experiment? The pandemic gave the world a chance to see what happens when daily life suddenly changes. Some of those changes, like working from home or using less transportation, might become long-term habits. Others may disappear as economies reopen. By analyzing which emission reductions lasted and which faded, we can start to imagine more realistic and effective ways to slow down climate change.

# **Data Exploration**

```{r}
library(tidyverse)
library(readr)
library(ggplot2)
library(tidyr)

data <- read_csv("data/owid-co2-data (1).csv")

```
## Narrowing Data to 2015-2022
```{r}
data_filtered <- data |>
  filter(year >= 2015, year <= 2022) |>
  filter(!is.na(iso_code) & nchar(iso_code) == 3)
```

```{r}
# Getting top 5 CO2 emitting countries by total GHG (excluding land use)
top_emitters <- data_filtered |>
  group_by(country) |>
  summarise(total_ghg = sum(total_ghg_excluding_lucf, na.rm = TRUE)) |>
  arrange(desc(total_ghg)) |>
  slice_head(n = 5) |>
  pull(country)

# Filtering data to include only those countries and select relevant variables
df <- data_filtered |>
  filter(country %in% top_emitters) |>
  select(country, year, ghg_excluding_lucf_per_capita,
         coal_co2, gas_co2, oil_co2, cement_co2, other_industry_co2)

```

```{r}
# Adding period category (pre, during, post COVID)
df <- df |>
  mutate(period = case_when(
    year <= 2019 ~ "pre_covid",
    year == 2020 ~ "during_covid",
    year >= 2021 ~ "post_covid"
  )) |>
  mutate(period = factor(period, levels = c("pre_covid", "during_covid", "post_covid")))

```

Per capita ghg emissions top 5 polluters
```{r}
ggplot(df, aes(x = factor(year), y = ghg_excluding_lucf_per_capita, color = country, group = country)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Per Capita GHG Emissions (Excl. Land Use) by Country (2015–2022)",
    x = "Year", y = "GHG per Capita (tCO₂e)"
  ) +
  theme_minimal()

```

```{r}
df_long <- df |>
  pivot_longer(cols = c(coal_co2, gas_co2, oil_co2, cement_co2, other_industry_co2),
               names_to = "sector", values_to = "emissions")

ggplot(df_long, aes(x = factor(year), y = emissions, color = sector, group = interaction(sector, country))) +
  geom_line() +
  geom_point() +
  facet_wrap(~ country) +
  labs(title = "Sector-specific CO₂ Emissions (2015–2022)", x = "Year", y = "Emissions (MtCO₂)") +
  theme_minimal()

```

```{r}
# Running ANOVA on each country if they have more than one period
for (c in unique(df$country)) {
  df_country <- df |> filter(country == c)
  if (n_distinct(df_country$period) > 1) {
    cat("\n--- ANOVA for", c, "---\n")
    model <- aov(ghg_excluding_lucf_per_capita ~ period, data = df_country)
    print(summary(model))
  } else {
    cat("\n--- Skipped ANOVA for", c, ": not enough periods ---\n")
  }
}

```

#### Data Description

We used data from Our World in Data’s greenhouse gas emissions dataset, which compiles country-level data from sources like the Global Carbon Project and BP Statistical Review [@owid-scaling-up-ai]. The dataset covers emissions from fossil fuels, industry, and land-use change.

For this project, we focused on the years 2015 to 2022 and selected the top five CO₂-polluting countries: China, the United States, India, Russia, and Japan. We used variables such as:

-   ghg_excluding_lucf_per_capita (our main variable),

-   Sector-based emissions: coal_co2, oil_co2, gas_co2, cement_co2, other_industry_co2.

#### Data Cleaning

We filtered the dataset to include:

-   Only rows with valid ISO country codes

-   Only years 2015–2022

-   The top 5 countries based on total GHG emissions (excluding land-use change)

We also created a period variable to group data into pre-COVID (2015–2019), during COVID (2020), and post-COVID (2021–2022). This helped us compare emissions trends across time. Minimal cleaning was needed beyond this filtering and labeling.


# Methods

Our reseach focuses on revealing relationships between countries and between predictor variables of greenhouse emissions per capita excluding land use. The dataset from Our World in Data includes emissions levels from the industrial revolution up to 2023 and is a robust dataset. The data are compiled from [sources]. The dataset is publicly posted to https://ourworldindata.org/co2-and-greenhouse-gas-emissions. To compare CO2 emissions before, during, and after the pandemic, we will be using the tidymodels package in R along with dplyr and [other packages]. We will make this comparison by comparing transportation and industrial emissions, then comparing total emissions from five key predictor categories including coal, oil, gas, cement and other. Our main predictor will be ghg_excluding_lucf_per_capita. We will be using modeltime to make a timeseries model against which predictor variables can be tested.

```{r, echo = TRUE}

#include a table of variables and their meaning (reference df above)
# copy metadata from orig proj to here
names(df)
```
#goal: find explainability of certain predictor variables

```{r, echo = TRUE}
#loading timeseries packages
library(tidyverse)
library(tsibble)
library(feasts)
library(lubridate)
library(modeltime)

#Make a date column from the year?
df <- df %>%  mutate(Date = as.Date(year), index = NULL)
#lubridate
ymd(df$year)
ymd(df$Date)

#format(Sys.Date(), '%Y.%m.%d')
#df <- df %>%  mutate(Date = Sys.Date())
#this didn't work; it just added date history
#I'm on the right track with a date column

#attaching a date string to the year to make the data work within the modeltime package
df <- df %>%  mutate(date_string = paste0(year, "-12-31"),
  Date = ymd(date_string))

#Check to see if modeltime requires a tsibble
# then, making a tsibble from cleaned data
df_tbl <- as_tsibble(df)
#from lecture 21
head(df_tbl)
```

## Making a recipe
``` {r, echo = TRUE}
#set a seed
set.seed(341)

#making a recipe based on timeseries data with relevant predictor variables
rec <-  recipe( ghg_excl_lucf_per_capita  ~ gas_co2 + cement_co2 + period + year + coal_co2 + oil_co2 + other_industry_co2) %>%
  # Log transform the predictor variables (aridity and p_mean)
  step_log(all_predictors()) %>%
  # Add an interaction term between aridity and p_mean
  step_interact(terms = ~ aridity:p_mean) |> 
  # Drop any rows with missing values in the pred
  step_naomit(all_predictors(), all_outcomes())

```

### Histograms of the data for normality

```{r, echo = TRUE}
#make a modeltime table to make a modeltime prediction

#potentially root mean squared error or modeltime_accuracy()

```


```{r, echo = TRUE}
#test the data for trends

#test the data for noise (lecture 21)


```

```{r, echo = TRUE}
# making a timeseries recipe


```

```{r, echo = TRUE}
# looking for significant difference in emissions per capita excluding land use change before, during, and after covid


```

```{r, echo = TRUE}
# looking for significant difference in transportation emissions between countries


#looking for signfiicant difference in industrial emissions between countries

```

```{r, echo = TRUE}
compare per capita most emissions to countries with highest overall emissions

```

```{r, echo = TRUE}
#visualize noise

#smooth to reduce noise

```

```{r, echo = TRUE}



```

```{r, echo = TRUE}



```
