---
title: "ESS 330 Final Project: Emissions in the Top CO₂ Polluting Countries Before and After COVID-19"
author:
  - name: "Yazeed Aljohani"
    affiliation: "Colorado State University"
  - name: "Josh Puyear"
    affiliation: "Colorado State University"
  - name: "Cade Vanek"
    affiliation: "Colorado State University"
date: 2025-04-24"
bibliography: references.bib
keywords: [CO2 emissions, COVID-19, GHG, workflow, ANOVA, sustainability]
abstract: |
  This project investigates trends in per capita greenhouse gas (GHG) emissions from the five largest CO2-polluting countries before, during, and after the COVID-19 pandemic. Using `ghg_excluding_lucf_per_capita` as the main [response?] variable, we analyze patterns in emissions and identify key predictors across sectors (e.g., manufacturing, transport). The study applies workflow modeling and ANOVA to examine emission changes and variation across countries. Results will inform strategies to reduce emissions and guide sustainable policy efforts.
---

# Introduction, Background, and Motivation

Climate change remains one of the most urgent global challenges, driven in large part by human greenhouse gas emissions. Among these, carbon dioxide is the most significant. Carbon dioxide is released through many activities such as burning fossil fuels, industrial production, and land-use practices. Even though international efforts like the Paris Agreement have pushed countries to reduce emissions, many continue to show increasing trends. A small group of countries contributes the most emissions globally, both in total and per person. These major contributors include countries like China, the United States, and India [@friedlingstein2019global].

This project focuses on the five largest carbon dioxide polluting countries: China, the United States, India, Russia, and Japan. These countries are responsible for a significant share of global emissions and reflect a range of economic systems, energy sources, and industrial structures. To allow meaningful comparisons, we center our analysis on greenhouse gas emissions per person, using the variable ghg_excluding_lucf_per_capita. This variable removes land-use change from the emissions total, isolating emissions from industrial activity and fossil fuel use. By using this approach, we can better understand how much pollution is tied to each person’s activity and consumption within each country.

The COVID-19 pandemic created a global disruption that affected travel, trade, and production. Lockdowns and other restrictions slowed down factories, reduced transportation, and caused shifts in energy use. As a result, global carbon dioxide emissions in 2020 dropped by approximately 5.4 percent, marking the largest single-year decrease in history [@forster2020current]. This sudden drop was mostly due to temporary changes. As countries began reopening in 2021 and 2022, emissions started climbing again.

Our goal is to analyze how emissions changed in these five countries before, during, and after the pandemic. We look at three main periods: before COVID from 2015 to 2019, the year COVID began in 2020, and the early recovery years from 2021 to 2022. By studying per person emissions during these periods, we can see whether there were real shifts in emissions behavior and whether those shifts continued into recovery. We also want to understand what kinds of emissions changed the most. For example, did transportation-related emissions drop more than industrial ones? Did countries with manufacturing-based economies recover differently than those with service-based ones?

To explore these questions, we use two main methods. First, we build a workflow model using the tidymodels package in R. This lets us see which emission sources are the strongest predictors of overall greenhouse gas output per person. For example, we compare emissions from coal, oil, gas, cement, and other industry sectors. We also include factors like population and economic output. By analyzing these predictors, we can understand which sources matter most in each country and how they are connected to changes over time. Second, we use ANOVA to test whether the changes in emissions across the three time periods are statistically meaningful. We run this analysis for each country and compare their patterns.

The reason for focusing on per person emissions is that total emissions alone do not tell the full story. Countries with large populations may seem like major polluters in total, but their emissions per person might be lower than smaller, high-emitting countries. A per person view also reflects differences in daily life, energy use, and infrastructure. This is important for climate policy, because it shows whether emissions are driven by large populations, high-intensity consumption, or both.

Our analysis helps link data analysis to climate solutions. By identifying what types of emissions dropped the most during the pandemic and which ones came back quickly, we can suggest where to focus efforts. If, for instance, oil-related emissions dropped during lockdowns but then returned quickly, that might show a need for deeper investment in alternative transportation options. If manufacturing emissions remained stable, that could show how industrial processes are harder to change without long-term plans and policies.

This project is also a chance to make emissions data more relatable. People often hear that emissions went down during the pandemic, but they may not know what caused the change or how long it lasted. By using clear visuals and country comparisons, we can help explain these trends to wider audiences. This is especially important as the world faces increasing climate-related disasters. Understanding emission trends can support smarter planning and better communication around sustainable transitions.

Lastly, we are motivated by a bigger question: what can we learn from this accidental global experiment? The pandemic gave the world a chance to see what happens when daily life suddenly changes. Some of those changes, like working from home or using less transportation, might become long-term habits. Others may disappear as economies reopen. By analyzing which emission reductions lasted and which faded, we can start to imagine more realistic and effective ways to slow down climate change.

# **Data Exploration**

```{r}
library(tidyverse)
library(readr)
library(ggplot2)
library(tidyr)

data <- read_csv("data/owid-co2-data (1).csv")

meta <- read_csv("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/owid-co2-codebook (1).csv")

```
## Narrowing Data to 2015-2022
```{r}
data_filtered <- data |>
  filter(year >= 2015, year <= 2022) |>
  filter(!is.na(iso_code) & nchar(iso_code) == 3)
```

```{r}
# Getting top 5 CO2 emitting countries by total GHG (excluding land use)- This is total cumulative ghg emissions

top_emitters <- data_filtered |>
  group_by(country) |>
  summarise(total_ghg = sum(total_ghg_excluding_lucf, na.rm = TRUE)) |>
  arrange(desc(total_ghg)) |>
  slice_head(n = 5) |>
  pull(country)

# Filtering data to include only those countries and select relevant variables
df <- data_filtered |>
  filter(country %in% top_emitters) |>
  select(country, year, ghg_excluding_lucf_per_capita,
         coal_co2, gas_co2, oil_co2, cement_co2, other_industry_co2)

```

```{r}
# Adding period category (pre, during, post COVID)
df <- df |>
  mutate(period = case_when(
    year <= 2019 ~ "pre_covid",
    year == 2020 ~ "during_covid",
    year >= 2021 ~ "post_covid"
  )) |>
  mutate(period = factor(period, levels = c("pre_covid", "during_covid", "post_covid")))

```

Per capita ghg emissions top 5 polluters
```{r}
ggplot(df, aes(x = factor(year), y = ghg_excluding_lucf_per_capita, color = country, group = country)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Per Capita GHG Emissions (Excl. Land Use) by Country (2015–2022)",
    x = "Year", y = "GHG per Capita (tCO₂e)"
  ) +
  theme_minimal()

```

```{r}
df_long <- df |>
  pivot_longer(cols = c(coal_co2, gas_co2, oil_co2, cement_co2, other_industry_co2),
               names_to = "sector", values_to = "emissions")

ggplot(df_long, aes(x = factor(year), y = emissions, color = sector, group = interaction(sector, country))) +
  geom_line() +
  geom_point() +
  facet_wrap(~ country) +
  labs(title = "Sector-specific CO₂ Emissions (2015–2022)", x = "Year", y = "Emissions (MtCO₂)") +
  theme_minimal()

```

```{r}
# Running ANOVA on each country if they have more than one period
for (c in unique(df$country)) {
  df_country <- df |> filter(country == c)
  if (n_distinct(df_country$period) > 1) {
    cat("\n--- ANOVA for", c, "---\n")
    model <- aov(ghg_excluding_lucf_per_capita ~ period, data = df_country)
    print(summary(model))
  } else {
    cat("\n--- Skipped ANOVA for", c, ": not enough periods ---\n")
  }
}

```

#### Data Description

We used data from Our World in Data’s greenhouse gas emissions dataset, which compiles country-level data from sources like the Global Carbon Project and BP Statistical Review [@owid-scaling-up-ai]. The dataset covers emissions from fossil fuels, industry, and land-use change.

For this project, we focused on the years 2015 to 2022 and selected the top five CO₂-polluting countries: China, the United States, India, Russia, and Japan. We used variables such as:

-   ghg_excluding_lucf_per_capita (our main variable),

-   Sector-based emissions: coal_co2, oil_co2, gas_co2, cement_co2, other_industry_co2.

#### Data Cleaning

We filtered the dataset to include:

-   Only rows with valid ISO country codes

-   Only years 2015–2022

-   The top 5 countries based on total GHG emissions (excluding land-use change)

We also created a period variable to group data into pre-COVID (2015–2019), during COVID (2020), and post-COVID (2021–2022). This helped us compare emissions trends across time. Minimal cleaning was needed beyond this filtering and labeling.


# Methods

Our reseach focuses on revealing relationships between countries and between predictor variables of greenhouse emissions per capita excluding land use. The dataset from Our World in Data includes emissions levels from the industrial revolution up to 2023 and is a robust dataset. The data are compiled from [sources]. The dataset is publicly posted to https://ourworldindata.org/co2-and-greenhouse-gas-emissions. To compare CO2 emissions before, during, and after the pandemic, we will be using the tidymodels package in R along with dplyr and [other packages]. We will make this comparison by comparing transportation and industrial emissions, then comparing total emissions from five key predictor categories including coal, oil, gas, cement and other. Our main predictor will be ghg_excluding_lucf_per_capita. We will be using modeltime to make a timeseries model against which predictor variables can be tested.

```{r, echo = TRUE}

#include a table of variables and their meaning (reference df above)
# copy metadata from orig proj to here
names(df)
```
# primary goal: make a linear model for energy consumption using lm

1. read in csvs of government responses to covid

loading libraries
```{r, echo = TRUE}
library(readr)
library(readxl)

```


```{r, echo = TRUE}
#loading all dataset
govres <- read_excel("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/Gov_Responses2Covid19_last.xlsx", sheet = "Dataset")

govresdes<- read_excel("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/Gov_Responses2Covid19_last.xlsx", sheet = "Description")

govres_econ <- read_excel("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/Gov_Responses_Sources.xlsx", sheet = "Econ sources")

govres_pubhealth <- read_excel("C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/data/Gov_Responses_Sources.xlsx", sheet = "Public health sources")

#View(govres)
#View(govres_source)
#view(govresdes)

```


1.5 colnames to figure out how columns are labeled

```{r, echo = TRUE}
names(govres)

summary(govres)

View(govres)

```


```{r, echo = TRUE}
# Getting top 5 CO2 emitting countries by total GHG (excluding land use)

#filter to top 5 countries

#need to actually just select for the five most ghg emitting countries
top_emitters_res <- govres |>
  group_by(country) |>
  filter(country %in% c("China", "India", "Russia", "United States of America", "Japan"))

# Filtering data to include only those countries and select relevant variables
df2 <- top_emitters_res %>% 
  select(country, date, Rigidity_Public_Health, Economic_Measures) %>% 
  mutate(Public = as.numeric(Rigidity_Public_Health)) %>% 
  mutate(Economic = as.numeric(Economic_Measures)) %>% 
  select(!c(Rigidity_Public_Health, Economic_Measures))
```

```{r, echo = TRUE}
df2_long <- df2 |>
  pivot_longer(cols = c(Public, Economic),
               names_to = "Restriction", values_to = "Rigidity") %>% 
  mutate(Date = as.Date(date))

#making a faceted plot
ggplot(df2_long, aes(x = Date, y = Rigidity, color = Restriction, group = interaction(Restriction, country))) +
    theme(
    axis.text.x = element_text(size = 6, angle = 60, hjust = 1),  # Remove extra comma here
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_x_date(date_labels = "%b %Y", 
               date_breaks = "3 months") +
  scale_y_continuous(limits = c(0, 1)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ country) +
  labs(title = "Restrictions During COVID's Lockdown", x = "Date", y = "Restriction Index") +
  theme_minimal()
 

ggsave("countryrestrict.pdf", plot = last_plot(), path = "C:/Users/Joshua Puyear/Documents/csu-undergrad/ess-330-joshp-2025/github/ESS-330-FinalProject-April23/docs", width = 12, height = 6)
```

Seeing if economic and public restrictions were statistically different
```{r, echo = TRUE}

library(car)

#shapiro test for Public restrictions
normspub <- df2 |> 
  nest(data = -country) |>
  mutate(
    Shapiro = map(data, ~ shapiro.test(na.omit(.x$Public))),  # Remove NAs before applying shapiro.test
    n = map_dbl(data, nrow),
    glance_shapiro = map(Shapiro, broom::glance)
  ) |>
  unnest(glance_shapiro)

flextable::flextable(dplyr::select(normspub, country, n, statistic, p.value)) |>
  flextable::set_caption("Shapiro-Wilk normality test for restrictions in each country")
#the public restriction dates of lockdown are not normally distributed

#testing economic restrictions
normsec <- df2 |> 
  nest(data = -country) |>
  mutate(
    Shapiro = map(data, ~ shapiro.test(na.omit(.x$Economic))),  # Remove NAs before applying shapiro.test
    n = map_dbl(data, nrow),
    glance_shapiro = map(Shapiro, broom::glance)
  ) |>
  unnest(glance_shapiro)

flextable::flextable(dplyr::select(normsec, country, n, statistic, p.value)) |>
  flextable::set_caption("Shapiro-Wilk normality test for restrictions in each country")
#economic restriction dates aren't normally distributed either
```

```{r, echo = TRUE}
#testing for statistical significance in differences between indices for how countries handled the lockdown

kruskal.test(Public ~ country, data = df2)

kruskal.test(Economic ~ country, data = df2)

```

post-hoc test to see which ones are different

```{r, echo = TRUE}
# using Dunn's test to see which variables are different
library(dunn.test)

# Perform Dunn's Test for pairwise comparisons
dunn.test(df2$Public, g = df2$country)

dunn.test(df2$Economic, g = df2$country)

```
Results:
The United States has significantly lower public restrictions than every other country except China. For further exploration, make a table of the number of days each type of lockdown took place in the US and China to compare. You would likely have to pivot df2. The difference in economic restriction indices was the same in each country.


2.1 look in colnames for energy consumption by sector or any indication. describe what energy metrics are present.

Proportion of CO2 emitted by top 5 countries in 2022

```{r, echo = TRUE}
#for all greenhouse gases

total_emissions <- data %>% 
  select(total_ghg, year, country, iso_code) %>%
  filter(!is.na(iso_code)) %>% 
  group_by(country, na.rm = TRUE) %>% 
  filter(year == 2022) %>% 
  ungroup()

#in the five most polluting countries
totem1 <- total_emissions %>% 
  filter(country %in% c("United States", "China", "India", "Japan", "Russia")) %>% 
  summarize(sum(total_ghg))

#in the rest of the world
totem2 <- total_emissions %>% 
  filter(!country %in% c("United States", "China", "India", "Japan", "Russia")) %>% 
  summarize(sum(total_ghg, na.rm = TRUE))

27168.36 / (27168.36 + 26138.06)

```
```{r, echo = TRUE}
#for co2_excluding_luc_per_capita

percap_emissions <- data %>% 
  select(ghg_excluding_lucf_per_capita, year, country, iso_code) %>%
  filter(!is.na(iso_code)) %>% 
  group_by(country, na.rm = TRUE) %>% 
  filter(year == 2022) %>% 
  ungroup()

#in the five most polluting countries
totemcap1 <- percap_emissions %>% 
  filter(country %in% c("United States", "China", "India", "Japan", "Russia")) %>% 
  summarize(mean(ghg_excluding_lucf_per_capita))

#in the rest of the world
totemcap2 <- percap_emissions %>% 
  filter(!country %in% c("United States", "China", "India", "Japan", "Russia")) %>% 
  summarize(mean(ghg_excluding_lucf_per_capita, na.rm = TRUE))
```

Results
The top five biggest co2 emitted of the world emitted a total of 27168.36 million metric tons of greenhouse gases in 2022. That year, the other 213 countries emitted metric tons of greenhouse gasses, meaning 50.97 percent of greenhouse gas emissions come from the top 5 most polluting countries. The mean emissions per capita in the top 5 biggest greenhouse gas emitters was 10.45 CO2 equivalents versus a mean of only 5.59 CO2 equivalents for the rest of the countries.

Making a faceted plot of carbon intensity of energy production for each of the 5 biggest ghg emitters

```{r, echo = TRUE}
#co2_per_unit_energy
 
df3 <- data_filtered %>% 
  group_by(country) %>% 
  filter(country %in% c("China", "India", "Russia", "United States", "Japan")) %>%
  select(country, year, co2_per_unit_energy,
         ghg_excluding_lucf_per_capita) %>% 
  mutate(percap = ghg_excluding_lucf_per_capita, energy = co2_per_unit_energy)


ggplot(df3, aes(x = factor(year), y = co2_per_unit_energy, color = country, group = country)) +
  geom_line() +
  geom_point() +
  labs(
    title = " 2015 - 2022 CO2 per unit energy emissions",
    subtitle = "Top 5 countries with greatest cumulative CO2 emissions",
    x = "Year", y = "CO2 produced per unit energy"
  ) +
  theme_minimal()


```

3.  make ANOVA for energy consumption for the top 5 highest ghg emitting countries and top 5 highest energy consumers to see if there's a significant difference in averages over time

```{r, echo = TRUE}
#shapiro test for Public restrictions
normspub2 <- df3 |> 
  nest(data = -country) |>
  mutate(
    Shapiro = map(data, ~ shapiro.test(na.omit(.x$energy))),  # Remove NAs before applying shapiro.test
    n = map_dbl(data, nrow),
    glance_shapiro = map(Shapiro, broom::glance)
  ) |>
  unnest(glance_shapiro)

flextable::flextable(dplyr::select(normspub2, country, n, statistic, p.value)) |>
  flextable::set_caption("Shapiro-Wilk normality test for energy consumption in each country")

#This table is giving normal distribution for the p-values, so we'll do ANOVA

```

ANOVA for co2 emissions per unit energy

```{r, echo = TRUE}
res_aov <- aov(formula = co2_per_unit_energy ~ country, data = df3)

#making sure residuals have a normal distribution
gghistogram(res_aov$residuals)

shapiro.test(res_aov$residuals)
#p-value is .8234, so the data fit a normal distribution
```


```{r, echo = TRUE}
library(ggpubr)
norm_usa <- df3 %>% group_by(country) %>% 
filter(country == "United States")

gghistogram(norm_usa$energy)

shapiro.test(df3$energy)

#it was strange how high the p-values were for the shapiro.test so I wanted to make a histogram. The USA histogram doesn't appear to have a normal distribution

```


p-values from ANOVA

```{r, echo = TRUE}
summary(aov(formula = co2_per_unit_energy ~ country, data = df3))
```
Results:

There is a statistically significant difference (p < 0.000) in CO2 per kilowatt-hour for the top 5 biggest cumulative CO2 emitters.


Look at lab 4 for ANOVA, edit code to reflect dataset

4.  energy consumption vs goverment responses to COVID linear model (is there a trend?)
look at lab 6 for lm and daily activity?

5. primary goal: make a ML model for energy consumption based on lab 6
government responses to COVID as a variable in ML model


## Predicting CO2 Emissions per unit Energy
``` {r, echo = TRUE}
#set a seed
set.seed(341)

#doing a correlation test for variables

data_filtered %>% 
cor.test(population + coal_co2_per_capita + oil_co2 + land_use_change_co2 + share_global_coal_co2 + share_global_luc + gdp)


```



#making a recipe to predict co2_per_unit_energy

rec <-  recipe(co2_per_unit_energy ~ gas_co2 + cement_co2 + period + year + coal_co2 + oil_co2 + other_industry_co2 ) %>%
  # Log transform the predictor variables (aridity and p_mean)
  step_log(all_predictors()) %>%
  # Add an interaction term between aridity and p_mean
  step_interact(terms = ~ aridity:p_mean) |> 
  # Drop any rows with missing values in the pred
  step_naomit(all_predictors(), all_outcomes())

```

### Histograms of the data for normality

```{r, echo = TRUE}

```


```{r, echo = TRUE}
#te
```

```{r, echo = TRUE}


```

```{r, echo = TRUE}

```

```{r, echo = TRUE}
# looking for significant difference in transportation emissions between countries


#looking for signfiicant difference in industrial emissions between countries

```

```{r, echo = TRUE}
#compare per capita most emissions to countries with highest overall emissions

```

```{r, echo = TRUE}
#visualize noise

#smooth to reduce noise

```

```{r, echo = TRUE}



```

```{r, echo = TRUE}



```
